---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Coursera: Course 5 - Project 1
==============================
**Author: Sharon Puddle**  

## Activity Monitoring Data

Data from a personal activity monitoring device has been collected over a two month period from an anonyous individual during the months of October and November 2012. It includes the number of steps taken in 5 minute intervals each day.

**Dataset:** [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52k]

### Variables
- **steps:** Number of steps taking in a 5-minute interval (missing values are coded as NA
- **date:** The date on which the measurement was taken in YYYY-MM-DD format
- **interval:** Identifier for the 5-minute interval in which measurement was taken  

## Loading and preprocessing the data
```{r setup, include=TRUE, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)

df <- read.csv("./data/activity.csv")
df <- tbl_df(df)

df_date <- df %>% filter(!is.na(steps)) %>% group_by(date) %>%
            summarize(total.steps = sum(steps),                       mean.steps = mean(steps))

df_int <- df %>% filter(!is.na(steps)) %>% group_by(interval) %>%
            summarize(total.steps = sum(steps), mean.steps=mean(steps))
```

## What is mean total number of steps taken per day?
```{r hist_total, echo=TRUE}
mean_steps <- as.integer(mean(df_date$total.steps))
median_steps <- median(df_date$total.steps)

ggplot(df_date, aes(x=total.steps)) + 
  geom_histogram(bins=12, colour="blue", fill="white") + 
  geom_vline(aes(xintercept=mean(total.steps, na.rm=T)), color="red", linetype="dashed", size=1) + 
  xlab("Total Steps") + ylab("Days") + 
  ggtitle("Total Steps Taken Each Day")
```
  
The mean and median total number of steps taken each day are:  
**Mean = `r mean_steps`**  
**Median = `r median_steps`**

## What is the average daily activity pattern?
```{r daily_pattern, echo=TRUE}
max_int <- df_int[which.max(df_int$mean.steps),1]

ggplot(data=df_int, aes(x=interval, y=mean.steps)) + 
  geom_line(colour="blue") + geom_point(colour="blue") +
  scale_x_continuous(breaks=seq(0,2400,200)) +
  xlab("5-Minute Intervals") + ylab("Average Steps") +
  ggtitle("Average Daily Activity Pattern")
```
  
The 5-minute interval which contains the maximum number of average steps is **`r max_int`**   

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
